---
title: "mhall-in-r"
author: "Matthew Henderson"
date: "15/02/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.path = "figure/",
  cache.path = "cache/",
  comment = "#>",
  collapse = TRUE,
  fig.width = 8
)
```


## Introduction

This repository contains code for generating latin squares
based on the theorem of Marshall Hall Jr. (1945) which states:

Given a rectangle of n - r rows and n columns such that
each of the numbers 1, 2, ..., n occurs in every row
and no number occurs twice in any column, then there exist
r rows which may be added to the given rectangle to form
a latin square.

In `R/add_rows.R` the function `add_rows` is implemented
which takes as input a latin rectangle (represented by a
data frame) and outputs a new latin rectangle with additional
rows (how many rows is also specified as input).

In this report we describe how `add_rows` function works
and demonstrate some of the generated outputs.

## Data frame representations of latin rectangles

A latin rectangle can be represented by an orthogonal
array and an orthogonal array looks just like a data frame
with variables for `row`, `column`, and `symbol`.

For example, the latin rectangle

```
1 2 3
2 3 1
```

can be represented by the orthgonal array data frame

```{r orthogonal_array_representation}
library(tidyverse)

tibble(
  row = c(rep(1, 3), rep(2, 3)),
  col = rep(1:3, 2),
  symbol = c(1, 2, 3, 2, 3, 1)
)
```

The `expand_grid` function makes it easier to construct
such a data frame from a vector of entries.

```{r expand_grid}
expand_grid(row = 1:2, column = 1:3) %>%
  mutate(symbol = c(1, 2, 3, 2, 3, 1))
```

A benefit this representation is that it makes it easy
to use the tidyverse packages. For example, we can plot
the latin square using `ggplot2`.

```{r first_plot, fig.height=2, fig.width=2}
expand_grid(row = 1:2, column = 1:3) %>%
  mutate(symbol = c(1, 2, 3, 2, 3, 1)) %>%
  ggplot(aes(column, row)) +
  geom_tile(aes(fill = symbol)) +
  geom_text(aes(label = symbol), colour = "white") +
  scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  theme(
    legend.position  = "none",
  )
```

## Extending latin rectangles

The `add_rows` function can take a data frame like this
as input and output an embedding of the latin rectangle
in a latin square.

```{r using_add_rows}
library(igraph)
library(tidygraph)

source("R/add_rows.R")
source("R/edge_tbl.R")
source("R/next_row.R")
source("R/to_tidygraph.R")

expand_grid(row = 1:2, column = 1:3) %>%
  mutate(symbol = c(1, 2, 3, 2, 3, 1)) %>%
  add_rows(3:3)
```
We can extend and plot all in one pipeline.

```{r extend_and_plot_pipeline, fig.height=2, fig.width=2}
expand_grid(row = 1:2, column = 1:3) %>%
  mutate(symbol = c(1, 2, 3, 2, 3, 1)) %>%
  add_rows(3:3) %>%
  ggplot(aes(column, row)) +
  geom_tile(aes(fill = symbol)) +
  geom_text(aes(label = symbol), colour = "white") +
  scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  theme(
    legend.position  = "none",
  )
```

## Existence of latin squares

It was understood early on that Marshall Hall Jr.'s theorem
provides an existence proof for latin squares. We can always
construct a 1 x n latin rectangle (it's just a permutation
of 1, ..., n). By Marhsall Hall Jr's theorem such a latin
rectangle can always be completed to a latin square, therefore
a latin square of order n exists.

For example, to construct a latin square of order 12.

```{r existence_plot}
expand_grid(row = 1:1, column = 1:12) %>%
  mutate(symbol = 1:12) %>%
  add_rows(2:12) %>%
  ggplot(aes(column, row)) +
  geom_tile(aes(fill = symbol)) +
  geom_text(aes(label = symbol), colour = "white") +
  scale_y_reverse() +
  coord_fixed() +
  theme_void() +
  theme(
    legend.position  = "none",
  )
```











